Apache Flink\texttrademark: Unified Stream and Batch Processing \\ in a Single Engine

### 2015-11-13 architecture.tex
Incorporated latest suggestions from doc

\para{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs. A Job Graph is a DAG of operators and connections between operators. There are two ``core'' APIs in Flink: the DataSet API for processing finite data sets (often referred to as ``batch processing''), and the DataStream API for processing potentially unbounded data streams (often referred to as ``stream processing''). Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create programs (Job Graphs) accepted by this engine. On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs. Currently these are the following:
\begin{inparaenum}[i)]
  \item Gelly, an API and library for processing graphs,
  \item FlinkML, an API and library for composing Machine Learning pipelines and
  \item Table, an API similar in spirit to Microsoft's LINQ.
\end{inparaenum}
A Flink cluster comprises of three types of processes: the client, the JobManager, and the TaskManager. The client takes the program code (written in a mixture of Flink's APIs), transforms it to a Job Graph, and submits it to the JobManager. The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types. DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in~\autoref{sec:batch}).

The JobManager is Flink's master node. It coordinates all message-passing during job execution by sending heartbeats to the TaskManagers, receiving statistics, controls the tasks' lifecycle,  coordinates the fault tolerance machanisms. The actual data processing takes place in the TaskManagers, or worker processes that they execute several tasks in multiple threads, and maintain data structures shared by all tasks (e.g., buffer pools) executed by a TM. TMs communicate directly with each other using a multiplexed TCP connection per TM pair.

>>>>

\para{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs. A Flink runtime program is a DAG of stateful operators connected with data streams. There are two core APIs in Flink: the DataSet API for processing finite data sets (often referred to as “batch processing”), and the DataStream API for processing potentially unbounded data streams (often referred to as “stream processing”). Flink’s core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create runtime programs executable by the engine. As such, it serves as the common fabric to abstract both bounded (batch) and unbounded (stream) processing. On top of the core APIs, Flink bundles domain-specific libraries and APIs that generate DataSet and DataStream API programs (currently FlinkML, Gelly, and Table).

A Flink cluster comprises of three types of processes: the client, the JobManager (JM - master), and one or more TaskManagers (RM - workers). The client takes the program code, transforms it to a runtime executable, and submits it to the JobManager. The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types. DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in ???).

The JobManager is Flink’s master node. It coordinates all message-passing during job execution by sending heartbeats to the TaskManagers and receiving statistics, controls the tasks’ lifecycle, and coordinates checkpointing used for recovery. The actual data processing takes place in the TaskManagers that they execute several tasks in multiple threads, and maintain data structures shared by all tasks (e.g., buffer pools) executed by a TM. TMs communicate directly with each other using a multiplexed TCP connection per TM pair.

\begin{figure}
\centering
	\includegraphics[width=.5\textwidth]{figs/architecture}
\caption{Flink's Process Model.}
\end{figure}

### 2015-11-13 batch.tex
Incorporated latest suggestions from doc

\section{Batch analytics on top of dataflows}
\label{sec:batch}
A bounded data set is a special case of an unbounded data stream. Thus, a streaming program that inserts all of its input data in a window can form a ``batch`` program and ``batch analytics``, or ``batch processing`` should be fully covered by Flink's features that we presented above. However, \begin{inparaenum}[i)]
  \item the syntax, i.e., the API for batch computation can be simplified (not need for window definitions, simpler joins and loops),
  \item we can simplify the fault tolerance mechanisms and
  \item we can apply query optimization borrowing ideas from MPP database systems.
\end{inparaenum}
For these reasons, Flink treats specially batch computations, by implementing the above optimizations. For enabling batch computations on top of it's streaming runtime, Flink embeds blocking versions of its operators (sorts, joins, etc) within its runtime. These operators simply block until they have received all of their input. Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad.~\cite{isard2007dryad}

With these established, we look that the batch-specific optimizations mentioned above: query optimization, and query processing on paged (managed) memory.

\para{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties. However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box [black-boxes], since the operators hide their semantics from the optimizer. For the same reason, cardinality and cost estimation methods are equally difficult to employ. Flink's optimizer employs a number of novel methods for overcoming these issues~\cite{blackBoxes, stratosphere, DBLP:journals/pvldb/EwenTKM12} for which we provide a short overview below. Flink's runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations. Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation~\cite{scopeOptimizer}, using a cost-based approach to choose among multiple physical plans. The cost includes network/disk I/O and CPU cost. To overcome the cardinality estimation issues that were mentioned earlier, Flink's optimizer uses hints that are provided by the programmer.

\para{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments. These memory segments resemble database blocks into which, Java objects representing the tuples that go through the runtime are serialized. Operations such as sorting, and joining, operate as much as possible on the binary data, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed. To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.  By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and implement cache-efficient and robust algorithms that scale gracefully in under memory constraints.

\para{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine. Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG~\cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program. The approach, implemented in Naiad~\cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations. A third approach was to design specialized engines around iterative processing along (e.g., Apache Giraph, and GraphLab) allowing to reduce the number of computations in each iteration.~\cite{low2012distributed}

Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory. This approach simulates a cyclic dataflow within a DAG engine making Flink competitive with specialized graph engines [], while outperforming the driver-based approach []. Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API. Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.

>>>>

\section{Batch analytics on top of dataflows}
\label{sec:batch}

A bounded data set is a special case of an unbounded data stream. Thus, a streaming program that inserts all of its input data in a window can form a “batch” program and “batch processing” should be fully covered by Flink’s features that we presented above. However, i) the syntax, i.e., the API for batch computation can be simplified (e.g., there is not need for artificial global window definitions, simpler join syntax), and iii) programs that process bounded data sets are amenable to additional optimizations, more efficient book keeping for fault tolerance, and staged scheduling. .

\vspace{2mm}
\noindent Flink approaches batch processing as follows:
\begin{itemize}
	\item Batch computations are executed by the same runtime as streaming computations. The runtime executable may be parameterized with blocked data streams to break up large computations in isolated stages that are scheduled successively.
	\item Periodic snapshotting is turned off. Instead, fault recovery is achieved by replaying the lost stream partitions from the latest buffered stream (possibly the source).
	\item Blocking operators (e.g., joins and sorts) are simply operator implementations that happen to block until they have consumed their entire input. The runtime is not aware of whether an operator is blocking or not. These operators use managed memory provided by Flink (either on- or off- the JVM heap), and can spill to disk if their inputs exceed their memory bounds.
	\item A dedicated DataSet API provides familiar abstractions for batch computations, namely a bounded fault-tolerant DataSet data structure, and transformations on DataSets (e.g., joins, aggregations, iterations, etc)
	\item A query optimization layer transforms a DataSet program to an efficient executable.
\end{itemize}

\para{Query Optimization.} Flink’s optimizer builds on techniques from parallel database systems such as plan equivalence, cost modeling and interesting property propagation. However, the arbitrary UDF-heavy DAGs that make up Flink’s dataflow programs do not allow a traditional optimizer to employ database techniques out of the box \cite{blackBoxes}, since the operators hide their semantics from the optimizer. For the same reason, cardinality and cost estimation methods are equally difficult to employ. Flink’s runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations. Flink’s optimizer enumerates different physical plans based on the concept of interesting properties propagation \cite{scopeOptimizer}, using a cost-based approach to choose among multiple physical plans. The cost includes network/disk I/O and CPU cost. To overcome the cardinality estimation issues that were mentioned earlier,  Flink’s optimizer uses hints that are provided by the programmer.

\para{Memory Management.} Building on database technology, Flink, instead of allocating objects in the JVM’s heap to represent buffered in-flight data records, serializes data into a  memory segments. Operations such as sorting and joining operate as much as possible on the binary data directly, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed. To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.  By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and use cache-efficient and robust algorithms [Graefe join paper] that scale gracefully under memory pressure.

\para{Native Iterations on top of Dataflows.}  Earlier dataflow approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG \cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program. The approach, implemented in Naiad \cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations. A third approach was to design specialized engines around iterative processing along (e.g., Giraph, and GraphLab) allowing to reduce the number of computations in each iteration \cite{low2012distributed}.

Flink follows an approach that retains the DAG-based runtime, but allows for special “head” and “tail” tasks to signify the beginning and end of iteration, that exchange data via shared memory. This approach simulates a cyclic dataflow within a DAG engine making Flink i ) competitive with specialized graph engines \cite{DBLP:journals/pvldb/EwenTKM12}, and outperforms the driver-based approach. Flink supports delta iterations, which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink’s Graph API. Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.

### 2015-11-13 intro.tex
Incorporated latest suggestions from doc

\section{Introduction}
\label{sec:intro}
Data stream processing (complex event processing / continuous queries) and static data processing (batch processing, OLAP) were traditionally considered as two very different types of applications. They were programmed using different programming models and APIs, and were executed by different systems (for example, dedicated streaming systems such as  Apache Storm, IBM Infosphere Streams, Microsoft Streaminsight, or Streambase versus relational databases or Apache Hadoop execution engines). Traditionally, batch data analysis made up for the lion share of the use cases, data sizes, and market, while streaming data analysis existed mostly in specialized applications.

It is becoming more and more apparent, however, that a huge number of today's large scale data processing use cases handle data that is, in reality, continuously produced over time. These continuous streams of data come for example from web logs, application logs, sensors, or as changes to application state in databases (transaction log records). Rather than treating the streams as streams, today's setups ignore the continuous and timely nature of the data production. Instead, data records are (often artificially) batched into static data sets (for example hourly/daily/monthly chunks) and then processed in a time-agnostic fashion. Data collection tools, workflow managers, and schedulers orchestrate the creation and processing of batches, in what is actually a continuous data processing pipeline. Architectural patterns like the "lambda architecture"~\cite{marz2015big} combine batch and stream processing systems to implement multiple paths of computation: A streaming fast path for timely approximate results, and a batch offline path for late accurate results.

@@

\noindent The contributions of this paper are the following:
\begin{itemize}
	\item We make the case for a unified architecture of stream and batch data processing, including specific optimizations that are only relevant for static data sets.
	\vspace{-3mm}
	\item We show how to make these dataflows fault tolerant via an asynchronous snapshotting mechanism (\autoref{sec:snapshots}).
	\vspace{-3mm}
	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (\autoref{sec:windows}), as well as a full-fledged batch processor (\autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
\end{itemize}

>>>>

\section{Introduction}
\label{sec:intro}
Data stream processing (e.g, as exemplified by Complex Event Processing systems) and static (batch) data processing (e.g., as exemplified by MPP databases and Hadoop ) were traditionally considered as two very different types of applications. They were programmed using different programming models and APIs, and were executed by different systems (for example, dedicated streaming systems such as  Apache Storm, IBM Infosphere Streams, Microsoft Streaminsight, or Streambase versus relational databases or execution engines for Hadoop, including Apache Spark, Apache Drill, etc). Traditionally, batch data analysis made up for the lion share of the use cases, data sizes, and market, while streaming data analysis mostly served specialized applications.

It is becoming more and more apparent, however, that a huge number of today's large scale data processing use cases handle data that is, in reality, continuously produced over time. These continuous streams of data come for example from web logs, application logs, sensors, or as changes to application state in databases (transaction log records). Rather than treating the streams as streams, today's setups ignore the continuous and timely nature of data production. Instead, data records are (often artificially) batched into static data sets (for example hourly/daily/monthly chunks) and then processed in a time-agnostic fashion. Data collection tools, workflow managers, and schedulers orchestrate the creation and processing of batches, in what is actually a continuous data processing pipeline. Architectural patterns like the "lambda architecture" \cite{marz2015big} combine batch and stream processing systems to implement multiple paths of computation: A streaming fast path for timely approximate results, and a batch offline path for late accurate results. All these approaches suffer from high latency (imposed by batches), high complexity (connecting and orchestrating several systems, and implementing business logic twice), as well as arbitrary inaccuracy, as the time dimension is not explicitly handled by the application code.

@@

\noindent The contributions of this paper are the following:
\begin{itemize}
	\item We make the case for a unified architecture of stream and batch data processing, including specific optimizations that are only relevant for static data sets.
	\vspace{-3mm}
	\item We show how streaming, batch, iterative, and interactive analytics can be represented as fault tolerant streaming dataflows (\autoref{sec:snapshots}).
	\vspace{-3mm}
	\item We discuss how we can build a full-fledged stream analytics system with a flexible windowing mechanism (\autoref{sec:windows}), as well as a full-fledged batch processor (\autoref{sec:batch}) on top of these dataflows, by showing how streaming, batch, iterative, and interactive analytics can be represented as streaming dataflows.
\end{itemize}

### 2015-11-13 snapshots.tex
Links moved to footnote
Incorporated latest suggestions from doc

\section{Fault Tolerance via Asynchronous Snapshots}
\label{sec:snapshots}
Fault tolerance in Flink is achieved by taking a snapshot of the execution graph at regular intervals. When failure occurs the state of the execution is restored from the latest snapshot. Upon recovery, in order to guarantee exactly-once processing semantics, the records (events) consumed since the dryad snapshot are reprocessed in the same order at each respective source. Normally this possible by using persistent message queueing systems, e.g. Apache Kafka.~\cite{CUSTOM:web/Kafka}

Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS).~\cite{carbone2015lightweight} It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner. It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream. Similar to Chandy-Lamport distributed snapshots~\cite{chandy1985distributed}, the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum. ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further. Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records. Furthermore, ABS~\cite{carbone2015lightweight} supports snapshotting of cyclic graphs.  The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated. The following figure illustrates the snapshot process.

\begin{figure}[h!]
	\centering  	
  	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
  	\vspace{-4mm}
	\caption{Asynchronous Barrier Snapshotting.}
	\vspace{-2mm}
	\label{fig:FlinkStack}
\end{figure}

>>>>

\section{Fault Tolerance via Asynchronous Snapshots}
\label{sec:snapshots}
Before jumping to the higher layers of the system, we detail how the common fabric (Flink’s runtime) guarantees consistent results in the case of failures. Since Flink data streams are replayable, discarding partial results and re-producing the lost partitions of the resulting streams (which possibly may trigger partial re-computation down to the sources in the case of non-buffered streams) is enough to guarantee fault tolerance when the computation is finite \cite{DBLP:conf/hotcloud/ZahariaCFSS10}. This is the mechanism used for Flink DataSet (batch) API programs.

In stream processing (Flink’s DataStream API), the computation is possibly infinite, which makes the above approach not practical, as possibly months of computation will need to be replayed. To bound recovery time, Flink takes a snapshot of the state of operators, including the current position of the input streams at regular intervals.

The core challenge lies in taking a consistent snapshot of all parallel operators without stopping the topology. In essence, the snapshot of all operators should refer to the same logical time in the computation. The mechanism used in Flink is called Asynchronous Barrier Snapshotting, or ABS \cite{carbone2015lightweight}. Barriers are messages injected into the input streams that correspond to a logical time, and logically separate the stream to the part whose effects will be included in the current snapshot, and the part that will be snapshotted later.

\begin{figure}[ht]
	\centering  	
  	\includegraphics[width=.75\textwidth]{figs/snaps.pdf}
  	\vspace{-4mm}
	\caption{Asynchronous Barrier Snapshotting.}
	\vspace{-2mm}
	\label{fig:FlinkStack}
\end{figure}

An operator receives barriers from upstream and first performs an alignment phase, making sure that the barriers from all inputs have been received. Then, the operator writes its state (e.g., contents of a sliding window, or custom data structures) to durable storage (the storage backend can be an external system, e.g., HDFS). Once the state has been backed up, the operator forwards the barrier downstream. ABS bears resemblances to the seminal Chandy-Lamport algorithm for asynchronous distributed snapshots \cite{chandy1985distributed}. However, because of the DAG structure of a Flink program and the alignment phase, ABS does not need to checkpoint in-flight records, guaranteeing that the data that needs to be written to reliable storage is kept to the theoretical minimum (i.e., only the current state of the operators).

Recovery from failures reverts the operator state to the most recently snapshotted one, and restarts the input streams from the latest barrier. This guarantees that the replayed work is limited to the processing between two checkpoints. We note that recovery need not start from the sources if intermediate streams have been partially buffered.

\vspace{2mm}
\noindent ABS provides several benefits:

\begin{itemize}
\item It guarantees exactly-once state updates without ever pausing the computation (e.g., as compared to discretized streams \cite{DBLP:conf/hotcloud/ZahariaCFSS10} which is essentially a synchronous implementation of snapshotting).
\item It is completely decoupled other forms of control messages, e.g., by events that trigger the computation of windows, thus not restricting the windowing mechanism to multiples of the checkpoint interval (again compare with \cite{DBLP:conf/hotcloud/ZahariaCFSS10})
\item It is completely decoupled from the mechanism used for reliable storage, allowing state to be backed up to file systems, databases, etc, depending on the larger environment in which Flink is used.
\end{itemize}
